---
title: "Bloodmeal Investigation"
output: html_notebook
---

# Introduction
I want to write down a way to sample the number of bites for each person
in each location and check that it looks right. In particular, let's write
down the version that draws from a negative binomial for the number of
bites per person, because that's the harder case.

# Distributions
Before I use the negative binomial distribution in R, I'll check that
R's definition matches something known. There are always alternative forms
of distributions, so it helps to be sure from the start.

According to R's documentation, the R negative binomial distribution has density
$$
 f(x;n, p) = \frac{\Gamma(x+n)}{\Gamma(n) x!} p^n (1-p)^x.
$$
Here, $\Gamma(n)=(n-1)!$ for integer $n$, so that's equivalent to
$$
 f(x;n, p) = \frac{(x+n-1)!}{(n-1)! x!} p^n (1-p)^x.
$$
For comparison, Wikipedia's [entry for the negative binomial](https://en.wikipedia.org/wiki/Negative_binomial_distribution)
writes it with a choose-notation.
$$
f(k; r,p) = \left(k+r-1\atop r-1\right)p^r(1-p)^k
$$
That's the same as above, swapping $(x,n,p)\rightarrow (k,r,p)$.

Fine. From this, we know the mean of R's distribution.
$$
\int_0^\infty xf(x;n,p)dx = \frac{n(1-p)}{p}
$$
The variance is the mean over $1-p$.
$$
\int_0^\infty (x-\mu)^2f(x;n,p)dx = \frac{n(1-p)}{p^2}
$$
# Parametrize the distribution
We want to parametrize the bloodmeal with two values, the EIR and the
dispersion. The mean should be the EIR.
$$
  E = \frac{n(1-p)}{p}
$$
When we use the term dispersion, we could measure that a number of different
ways. One is the standard deviation.
$$
D = \frac{\sqrt{n(1-p)}}{p}
$$
If I look at the C++ implementation, it seems to use a ratio of dispersion
to the mean.
$$
\frac{D}{E} = \frac{p}{1-p}
$$
Let's go with that, which implies a simple formulat for $p$.
$$
p = \frac{D}{D+E}
$$
Given this value for $p$, what $n$ should be.

We need the mean number of bites to equal the EIR.
$$
n = \frac{Ep}{1-p}
$$
That looks like the C++ function.
```{r cpp_bloodmeal, eval = FALSE}
int queue_bites_nbinom::sample_bites(const double eir){
  double p = disp / (disp + eir);
  int k = (eir*p)/(1.-p);
  return R::rnbinom(k,p);
}
```
It's the same in R.
```{r r_bloodmeal}
bloodmeal_nbinom <- function(eir, dispersion) {
  p <- dispersion / (dispersion + eir)
  k <- (eir * p) / (1 - p)
  rnbinom(1, size = k, prob = p)
}
```


# R sampling

There are humans, $h$, and locations $l$.
People have biting weights, $b_h$. The mosquitoes in a location have a biting
rate, $c_l$. We start with a list of which people dwelled in which locations.
Let's start by giving each person a home location and drawing their dwell time
from a multinomial over the locations.
```{r build_world}
l_cnt <- 2L
humans_per_location <- 3L
h_cnt <- humans_per_location * l_cnt
home <- rep(1:l_cnt, each = humans_per_location)
b <- runif(h_cnt, 0.5, 1)
c <- runif(l_cnt, 0.2, 0.5)
# A human in a location has a distribution for where they go.
travel_distribution <- function (l_idx, l_cnt) {
  shape = 1
  preference <- rexp(l_cnt, shape)
  highest <- which.max(preference)
  save <- preference[l_idx]
  preference[l_idx] <- highest
  preference[highest] <- save
  preference / sum(preference)
}
location_travel <- vapply(
  1:l_cnt,
  function(l_idx) travel_distribution(l_idx, l_cnt), FUN.VALUE = numeric(l_cnt)
  )
stopifnot(abs(colSums(location_travel) - rep(1, l_cnt)) < 1e-10)
# Each human's travel is a draw from that distribution, for 30min segments.
travel_pattern <- function(h_idx) {
  time_chunks <- rmultinom(1, 48, location_travel[, home[h_idx]])
  time_chunks / sum(time_chunks)
}
tar <- vapply(1:h_cnt, travel_pattern, numeric(l_cnt))
stopifnot(abs(colSums(tar) - rep(1, l_cnt)) < 1e-10)
stopifnot(dim(tar) == c(l_cnt, h_cnt))
```

Now turn that into a total biting weight per location and person.
```{r total_biting_weight}
tar_biting_weight <- tar %*% diag(b)
location_biting_weight <- rowSums(tar_biting_weight)
location_biting_weight
```
Let's give the location a tendency for mosquitoes. Make it rocky, so that
there is some chance of none, some of many.
```{r mosquitoes_at_location}
multitudes_of_mosquitoes <- rbinom(l_cnt, 5, 0.4)
mosquitoes <- rbinom(l_cnt, 1000*multitudes_of_mosquitoes, 0.1)
```

We could sample, for a location, the number of bites. That would look like
this.
```{r mosquito_bites, eval = FALSE}
bites_at_location <- rbinom(l_cnt, mosquitoes, c)
```
But how would we assign them to people?

Instead, take the biting rate and turn that back onto each person.
```{r}
biterate_location <- mosquitoes * c
print(biterate_location)
fraction_of_bites_to_each_human <- diag(1 / rowSums(tar_biting_weight)) %*% tar_biting_weight
fraction_of_bites_to_each_human
average_bites_per_human <- biterate_location %*% fraction_of_bites_to_each_human
print(average_bites_per_human)
print(fraction_of_bites_to_each_human * biterate_location)
```
We can use that as our average for sampling.
```{r}
dispersion <- 1.5
bites_per_human <- vapply(
  average_bites_per_human,
  function(bites) bloodmeal_nbinom(bites, dispersion),
  numeric(1)
  )
```
Then allocate these back to the locations with a multinomial.
```{r allocate_bites_back}
bites_matrix <- vapply(
  1:h_cnt,
  function(h_idx) rmultinom(1, bites_per_human[h_idx], tar_biting_weight[, h_idx]),
  numeric(l_cnt)
)
bites_at_location <- rowSums(bites_matrix)
stopifnot(sum(bites_at_location) == sum(bites_per_human))
```

# Distribution from that sampling

We worked out a way to sample. Now let's run this a bunch of times and see if
the distribution of samples looks OK. I took the above code and put it
into `bloodmeal_daily.R` so that we have functions to call.

```{r}
world <- build_world(5, 20)
travel <- sample_travel(world)
mosquitoes <- sample_mosquitoes(world)
```
## Some debugging on the code version
```{r}
b = world$b
c = world$c
dispersion = world$dispersion
h_cnt <- world$h_cnt
l_cnt <- world$l_cnt

bite_weight.lh <- travel %*% diag(b)
bite_weight.l <- rowSums(bite_weight.lh)

bite_rate.l <- mosquitoes * c
fraction_of_bites_to_each_human <- diag(1 / bite_weight.l) %*% bite_weight.lh
bite_rate.h <- bite_rate.l %*% fraction_of_bites_to_each_human
bite_rate.lh <- fraction_of_bites_to_each_human * bite_rate.l

bites.h <- vapply(
  bite_rate.h,
  function(bites) bloodmeal_nbinom(bites, dispersion),
  numeric(1)
)
bites.lh <- vapply(
  1:h_cnt,
  function(h_idx) {
    rmultinom(1, size = bites.h[h_idx], prob = bite_rate.lh[, h_idx])
    },
  numeric(l_cnt)
)
```

```{r}
sample_cnt <- 10000
bites_l_draws <- vapply(1:sample_cnt, function(i) {
    bites <- sample_bites(travel, mosquitoes, world)
    rowSums(bites)
  },
  numeric(world$l_cnt))
hist(bites_l_draws[4, ])
```
This KDE uses jitter on the integer values. It looks pretty bad, though.
```{r, eval = FALSE}
for (loc_idx in 1:l_cnt) {
  loc1 <- kde1d::kde1d(bites_l_draws[loc_idx,], xmin = 0, deg = 2)
  if (loc_idx == 1) {
    plot(loc1)
  } else {
    lines(loc1)
  }
}
```

```{r}
expected_location_bites <- mosquitoes * world$c
trimmed_mean <- function(data, indices) {
  mean(data[indices], trim = 0.2)
}
full_mean <- function(data, indices) {
  mean(data[indices])
}
mean_absolute_deviation <- function(data, indices) {
  max(data[indices], constant = 1.4826)
}
compare_over_statistic <- function(statistic) {
  # Generate a robust estimator of the Poisson distribution
  mean_generated <- lapply(1:world$l_cnt, function(l_idx) {
    basic_rate <- rpois(1000, expected_location_bites[l_idx])
    boot::boot(
      data = basic_rate,
      statistic = statistic,
      R = 599
    )
  })
  # Compare with same robust estimator on observed values
  mean_est <- lapply(1:world$l_cnt, function(l_idx) {
      boot::boot(
      data = bites_l_draws[l_idx,],
      statistic = statistic,
      R = 599
      )
    })
  sampled_bites <- vapply(mean_est, function(m) m$t0, numeric(1))
  expected_bites <- vapply(mean_generated, function(m) m$t0, numeric(1))
  cbind(sampled_bites, expected_bites)
}
print(compare_over_statistic(trimmed_mean))
compare_over_statistic(mean_absolute_deviation)
```
The net result is that I'm concerned about the final distributions
because they don't appear to be Poisson in variance.

Go back and check the bloodmeal. It has the correct expected value.
```{r}
expected_eir <- 10
bloodsamples <- vapply(
  1:10000,
  function(x) bloodmeal_nbinom(expected_eir, 3),
  numeric(1)
)
print(c(mean(bloodsamples), expected_eir))
hist(bloodsamples)
```

