---
title: "Hook Some Things Together"
output: html_notebook
---

```{r load_libs, show = FALSE}
library(macro)
library(data.table)
```


# Introduction

We have some components that should form a malaria simulation. Let's walk through creating them and hooking them together.

There are four parts: a human health module, a human movement module, a mosquito module, and a bloodmeal module. Each needs to be created to describe a set of locations with humans and mosquitoes. We can start with a general notion, for instance that there are three locations with 40 humans among them.
```{r initial_sense}
parameters <- list(
  human_cnt = 40L,
  location_cnt = 3L,
  duration_days = 10
)
parameters$humans.l <- c(rmultinom(1, parameters$human_cnt, c(0.1, 0.3, 0.6)))
```
Here, the `.l` means it's a vector over locations. We'll use `.h` to mean a vector over humans. It's a convenient signal about what a variable's dimensions are for.


# Movement Module

This module is independent of the others, so let's take care of it first. It needs a stochastic matrix for movement, where rows are people and columns are locations.

```{r make_simpletrip}
start_at_home <- rep(1:parameters$location_cnt, parameters$humans.l)
stopifnot(length(start_at_home) == parameters$human_cnt)

stochastic_matrix <- matrix(runif(parameters$location_cnt^2), nrow = parameters$location_cnt)
diag(stochastic_matrix) <- 0
# The rows should sum to one.
stochastic_matrix <- diag(1 / rowSums(stochastic_matrix)) %*% stochastic_matrix
return_matrix <- matrix(1, nrow = parameters$location_cnt, ncol = parameters$location_cnt)
diag(return_matrix) <- 0
return_matrix <- diag(1 / rowSums(return_matrix)) %*% return_matrix

simpletrip_parameters <- list2env(list(
  trip_rate = rep(0.1, parameters$location_cnt),
  trip_dest = stochastic_matrix,
  return_home_rate = return_matrix,
  npatch = parameters$location_cnt,
  home = start_at_home,
  current = start_at_home
))
movement <- simple_trip_module(simpletrip_parameters)
movement <- mash_step(movement, parameters$duration_days, NULL)
locations <- person_path(movement)
```

```{r}
traj <- movement$trajectory[1:movement$trajectory_cnt]
whos <- vapply(traj, function(entry) entry$id, FUN.VALUE=integer(1))
npeople <- parameters$human_cnt
max_moves <- max(vapply(1:npeople, function(i) sum(whos == i), FUN.VALUE=numeric(1)))

move_dt <- data.table(
  ID = movement$state$who,
  Start = movement$state$current
  )
move_dt[, paste0(c("Time", "Location"), rep(1:max_moves, each = 2))] <- 0L
move_dt[, paste0(c("Time", "Location"), rep(1:max_moves, each = 2))] <- NA
event <- movement$trajectory[[1]]
h_idx <- 1
move_idx <- which(is.na(move_dt[h_idx,]))[1]
if (move_idx == 3) {
  move_dt[h_idx, Start] <- event$prev_location
}  # No fixup if not the first event for this person.
move_dt[h_idx, move_idx] <- event$curtime
move_dt[h_idx, move_idx + 1] <- event$curr_location
```


```{r step_simpletrip}
locations
```
The resulting data structure is

* a list with an entry for each location.
* Each member is a data.table with three columns: `arrive`, `leave`, and `time`.
* The values in `arrive` or `leave` are the ID of the individual, where either one or the other is the ID, and the other one is `NaN`.


# Health Module

The parameters for the health module are few.
```{r make_health}
health_parameters <- list(
  recovery_rate = 1 / 200,
  people_cnt = parameters$human_cnt,
  duration_days = parameters$duration_days,
  initial_pfpr = 0.4
)
health <- forced_si_module(health_parameters)
current_time <- 0
bites <- lapply(
  1:parameters$human_cnt,
  function(h_idx) {
    times <- rexp(1, 0.1)
    sort(times[times < 10])
  }
  )
health <- mash_step(health, bites)
health_path <- human_disease_path(health)
```

```{r show_trajectory}
health_path
```


# Mosquito Module

```{r make_mosquito_module}
year_days <- 365
mosquito_parameters <- list(
  duration = parameters$duration_days,
  N = parameters$location_cnt,
  lambda = matrix(rep(0.1, year_days * parameters$location_cnt), nrow = parameters$location_cnt),
    psi = diag(parameters$location_cnt),  # diffusion matrix
    EIP = rep(12, year_days),  # Extrinsic incubation period,
    maxEIP = 12,  # Maximum length of EIP.
    p = 0.9,  # survival
    # human blood feeding rate, the proportion of mosquitoes that feed on humans each day
    a = 0.6,
    year_day_start = 1,
    # Does not affect calculation value.
    adult_scale = NULL  # Scale for mosquito population.
  )
module <- mosquito_rm_module(mosquito_parameters)
bloodmeal_dt <- data.table(expand.grid(1:parameters$location_cnt, 1:parameters$duration_days))
names(bloodmeal_dt) <- c("Location", "Time")
bloodmeal_dt$Time <- bloodmeal_dt$Time - 1
bloodmeal_dt$Bites <- sample(10:20, nrow(bloodmeal_dt), replace = TRUE)
module <- mash_step(module, bloodmeal_dt)
bites_dt <- mosquito_path(module)
bites_dt
```

# Bloodmeal module

```{r make_bloodmeal}
params <- list(
  human_cnt = 8L,
  location_cnt = 5L,
  duration = 10,
  day_duration = 1,
  dispersion = 1.5,
  day_cnt = 10
)
health_dt <- sample_health_infection_status(params$human_cnt, params$duration)
bite_weight <- runif(params$human_cnt, 0.2, 0.8)
movement_dt <- sample_move_location(params$human_cnt, params$location_cnt, params$duration)
mosquito_dt <- sample_mosquito_myz(params$location_cnt, params$duration)

bloodmeal <- bloodmeal_density_module(params)
bloodmeal <- mash_step(bloodmeal, health_dt, movement_dt, mosquito_dt)
human_infections <- infects_human_path(bloodmeal)
mosquito_infections <- infects_mosquito_path(bloodmeal)
mosquito_infections
```

