---
title: "Hook Some Things Together"
output: html_notebook
---

```{r load_libs, show = FALSE}
library(macro)
library(data.table)
```


# Introduction

We have some components that should form a malaria simulation. Let's walk through creating them and hooking them together.

There are four parts: a human health module, a human movement module, a mosquito module, and a bloodmeal module. Each needs to be created to describe a set of locations with humans and mosquitoes. We can start with a general notion, for instance that there are three locations with 40 humans among them.
```{r initial_sense}
parameters <- list(
  human_cnt = 40L,
  location_cnt = 3L,
  duration_days = 10
)
parameters$humans.l <- c(rmultinom(1, parameters$human_cnt, c(0.1, 0.3, 0.6)))
```
Here, the `.l` means it's a vector over locations. We'll use `.h` to mean a vector over humans. It's a convenient signal about what a variable's dimensions are for.

# Movement Module

This module is independent of the others, so let's take care of it first. It needs a stochastic matrix for movement, where rows are people and columns are locations.

```{r make_simpletrip}
start_at_home <- rep(1:parameters$location_cnt, parameters$humans.l)
stopifnot(length(start_at_home) == parameters$human_cnt)

stochastic_matrix <- matrix(runif(parameters$location_cnt^2), nrow = parameters$location_cnt)
diag(stochastic_matrix) <- 0
# The rows should sum to one.
stochastic_matrix <- diag(1 / rowSums(stochastic_matrix)) %*% stochastic_matrix
return_matrix <- matrix(1, nrow = parameters$location_cnt, ncol = parameters$location_cnt)
diag(return_matrix) <- 0
return_matrix <- diag(1 / rowSums(return_matrix)) %*% return_matrix

simpletrip_parameters <- list2env(list(
  trip_rate = rep(0.1, parameters$location_cnt),
  trip_dest = stochastic_matrix,
  return_home_rate = return_matrix,
  npatch = parameters$location_cnt,
  home = start_at_home,
  current = start_at_home
))
movement <- simple_trip_module(simpletrip_parameters)
movement <- mash_step(movement, parameters$duration_days, NULL)
locations <- location_path(movement)
```
The resulting data structure is a list with an entry for each location. Each member is a data.table with three columns: `arrive`, `leave`, and `time`. The values in `arrive` or `leave` are the ID of the individual, where either one or the other is the ID, and the other one is `NaN`.
