---
title: "Make modules and run"
output: html_notebook
---

```{r load_libs, show = FALSE}
library(macro)
library(data.table)
library(ggplot2)
```

# Introduction

Very directly make the modules and run them. No testing here.

```{r initial_sense}
parameters <- list(
  human_cnt = 40L,
  location_cnt = 3L,
  duration_days = 10
)
# Count of humans in each location.
parameters$humans.l <- c(rmultinom(1, parameters$human_cnt, c(0.1, 0.3, 0.6)))
stopifnot(length(parameters$humans.l) == parameters$location_cnt)
```
Here, the `.l` means it's a vector over locations. We'll use `.h` to mean a vector over humans. It's a convenient signal about what a variable's dimensions are for.

# Movement Module

This module is independent of the others, so let's take care of it first. It needs a stochastic matrix for movement, where rows are people and columns are locations.

```{r make_simpletrip}
start_at_home <- rep(1:parameters$location_cnt, parameters$humans.l)
stopifnot(length(start_at_home) == parameters$human_cnt)

stochastic_matrix <- matrix(runif(parameters$location_cnt^2), nrow = parameters$location_cnt)
diag(stochastic_matrix) <- 0
# The rows should sum to one.
stochastic_matrix <- diag(1 / rowSums(stochastic_matrix)) %*% stochastic_matrix
return_matrix <- matrix(1, nrow = parameters$location_cnt, ncol = parameters$location_cnt)
diag(return_matrix) <- 0
return_matrix <- diag(1 / rowSums(return_matrix)) %*% return_matrix

simpletrip_parameters <- list2env(list(
  trip_rate = rep(0.1, parameters$location_cnt),
  trip_dest = stochastic_matrix,
  return_home_rate = return_matrix,
  npatch = parameters$location_cnt,
  home = start_at_home,
  current = start_at_home,
  duration_days = parameters$duration_days
))
movement <- simple_trip_module(simpletrip_parameters)
```


# Health Module

The parameters for the S-I health module are few.
```{r make_health}
health_parameters <- list(
  recovery_rate = 1 / 200,
  people_cnt = parameters$human_cnt,
  duration_days = parameters$duration_days,
  initial_pfpr = 0.2
)
health <- human_si_module(health_parameters)
```


# Mosquito Module

```{r make_mosquito_module}
year_days <- 365
mosquito_parameters <- list(
  duration = parameters$duration_days,
  N = parameters$location_cnt,
  lambda = matrix(
    rep(100, year_days * parameters$location_cnt),
    nrow = parameters$location_cnt),
  psi = diag(parameters$location_cnt),  # diffusion matrix
  biting_weight = 0.4,  # The human biting weight. Should be from human.XXX
  EIP = rep(12, year_days),  # Extrinsic incubation period,
  maxEIP = 12,  # Maximum length of EIP.
  p = 0.9,  # survival
  # human blood feeding rate, the proportion of mosquitoes that feed on humans each day
  a = 0.6,
  infected_fraction = rep(0.1, parameters$location_cnt),
  year_day_start = 1
)
mosquito_module <- mosquito_rm_module(mosquito_parameters)
```


# Bloodmeal module
```{r bloodmeal_parameters}
params <- list(
  human_cnt = parameters$human_cnt,
  location_cnt = parameters$location_cnt,
  duration = parameters$duration_days,
  day_duration = 1,
  dispersion = 1.5,
  day_cnt = 10,
  day_start = 1,
  biting_weight = 0.5
)
bloodmeal <- bloodmeal_density_module(params)
```

# Assemble and run

```{r}
macro:::local_logging("info")
modules = list(
  location = movement,
  health = health,
  bloodmeal = bloodmeal,
  mosquito = mosquito_module
)
step_cnt <- 5
observer <- complete_observer()
singleobs <- vector(mode = "list", length = step_cnt)
msgnames <- c(
  "location", "bloodmeal_human", "bloodmeal_mosquito", "health", "mosquito")
dump_if <- function(step, idx) { step == "bloodmeal" }
seen <- list()
for (main_idx in 1:step_cnt) {
  cat(paste("Running 10-day step", main_idx, "\n"))
  modules <- step_mainloop(modules, observer, main_idx, dump_if)
  singleobs[[main_idx]] <- lapply(
    msgnames,
    function(msgtype) {
      data.table::rbindlist(modules$observer[[msgtype]])
    }
  )
  names(singleobs[[main_idx]]) <- msgnames
}
observations <- lapply(
  msgnames,
  function(msgname) {
    data.table::rbindlist(
      lapply(singleobs, function(obs) obs[[msgname]])
    )
  }
)
names(observations) <- msgnames
```
```{r}
homeloc <- rep(1:parameters$location_cnt, parameters$humans.l)
state <- homeloc
loc_df <- observations$location
sumloc <- matrix(0, ncol = nrow(loc_df), nrow = (parameters$location_cnt + 1))
for (ridx in 1:nrow(loc_df)) {
  state[loc_df[ridx, ID]] <- loc_df[ridx, Location]
  sumloc[, ridx] <- c(
    loc_df[ridx, Time],
    hist(state, breaks = 0:parameters$location_cnt, plot = FALSE)$counts
  )
}
loc_dt <- data.table::data.table(t(sumloc))
# Keep only the last of any times that are identical.
loc_dt <- loc_dt[!duplicated(loc_dt[, 1], fromLast = TRUE)]
colnames(loc_dt) <- c("Time", paste0("Loc", 1:parameters$location_cnt))
ggplot(loc_dt, aes(Time, Loc1, colour = "green")) + geom_line()
  ggplot(loc_dt, aes(Time, Loc2, colour = "black")) + geom_line()
  ggplot(loc_dt, aes(Time, Loc3, colour = "blue")) + geom_line()

```


```{r}
bm <- observations$bloodmeal_mosquito
bm$Location <- as.factor(bm$Location)
ggplot(bm, aes(Time, Bites, colour = Location)) + geom_point() +
  labs(title = "Bites in Each Location During Time Step", y = "Bites per Day")
```

Bloodmeal's bites for the human.
```{r}
bh <- observations$bloodmeal_human
ggplot(bh, aes(times)) + geom_histogram(binwidth = 10 / 40) +
  labs(title = "Bites for all Humans During Time Step", y = "Bites per Quarter Day")
```


```{r}
bmosy <- observations$mosquito
bmosy1 <- bmosy[bmosy$Location == 1,]
ymax <- 1.1 * max(bmosy1$M, bmosy1$Y, bmosy1$Z)
plot(bmosy1$Time, bmosy1$M, type = "p", pch = 16,
     ylim = c(0, ymax))
points(bmosy1$Time, bmosy1$Y, col = "blue", pch = 16)
points(bmosy1$Time, bmosy1$Z, col = "green", pch = 16)
legend("bottomright", legend = c("M", "Y", "Z"), fill = c("black", "blue", "green"))
```
```{r}
library(ggplot2)
mosy <- observations$mosquito[, c("Location", "Time", "M", "Y", "Z")]
ggplot(data=mosy, aes(x=Time, y = Z)) + geom_point() + facet_wrap(~Location)
```
```{r}
# Number sick for each location.
state_dt <- data.table::data.table(
  home = rep(1:parameters$location_cnt, parameters$humans.l),
  state = numeric(length=parameters$human_cnt)
)
loc_df <- observations$location
h_df <- observations$health
sumh <- matrix(0, ncol = nrow(h_df), nrow = (parameters$location_cnt + 1))
for (ridx in 1:nrow(h_df)) {
  id <- h_df[ridx, ID]
  newlevel <- h_df[ridx, Level]
  state_dt[id, `:=`(state = newlevel)]
  sumh[, ridx] <- c(
    h_df[ridx, Time],
    state_dt[, sum(state), home][, V1]
  )
}
h_dt <- data.table::data.table(t(sumh))
# Keep only the last of any times that are identical.
h_dt <- h_dt[!duplicated(h_dt[, 1], fromLast = TRUE)]
colnames(h_dt) <- c("Time", paste0("Loc", 1:parameters$location_cnt))
ggplot(h_dt, aes(Time, Loc1, colour = "green")) + geom_line()
  ggplot(h_dt, aes(Time, Loc2, colour = "black")) + geom_line()
  ggplot(h_dt, aes(Time, Loc3, colour = "blue")) + geom_line()
```

