---
title: "Human Module with SI Dynamics"
output:
  html_notebook:
    df_print: paged
  pdf_document:
    toc: true
    df_print: kable
    number_sections: yes
header-includes:
  - \usepackage{palatino}
bibliography: pfSI.bib
---
```{r load_libraries, include = FALSE}
library(data.table)
```

# Overview

This notebook describes a MASH module for humans who can be
either susceptible or infectious. We call it `pfSI`.

There is an existing module in a former implementation of MASH,
so this document will walk through that implementation in order
to write out this one.

We ask here the questions we ask of any model, be it discrete, continuous,
deterministic or stochastic. What are the states,
what are the transitions, and how do you initialize the model?
The implementation given is continuous time, so it creates individual events,
which should make it easier to track what's happening.


# Problem Description

## State Space

The first thing to define for this model is the set of possible
states a human can be in. We split these into three categories.
Some properties of a person are unchanging, and we call them
_frailty._ Some properties of a person can be figured out. For instance,
if biting weight is a function of age, then you can figure it out from
the age without storing biting weight. We call these _predictable variables._
The rest is state.

* Frailty
  + birth date
  + mosquito-to-human transmission efficiency, _b_
  + human-to-mosquito transmission efficiency, _c_
* State
  + alive or dead - a boolean.
  + disease status - as susceptible, latent, infectious, or protected from infection
  + fever - reachable from the infectious state (or latent?)
  + PE vaccination - lowers biting weight by factor (1 - pe_block_pf)
* Predictable Variables
  + biting weight - this represents the amount of skin this person has,
    so it changes with age. Lowered by PE vaccination
  + kappa - if the person is infected and _not in a resevoir,_ this is c x (biting weight).
    Lowered by GS vaccination by a factor c = c x (1 - gs_block_pf).
  + EIR - the rate the person is getting bitten. For someone in a resevoir,
    this is the resevoir EIR. For someone not in a resevoir, this
    is max(0, beta x biting_weight / patch_biting_weight).
  + incidence - the number of infectious people in a patch. This is the sum of all infectious.


## Event Space

Here, variable names are sometimes shortened. Mean is mn, and variance is vr.
There is a notation for events that says what state you leave and what
you enter. The minus is leave and the + is enter. For instance, if a single
person leaves the latent stage and enters the infectious stage, the
notation is (-L, +I). If they have to be in a stage in order for something else
to run, you will see (-I, +F, +I), meaning fever can't happen unless the person
is infected, but they stay infectious.

* Infection
  + infection - from state L to I (-L, +I), exponential with rate duration
  + clearance - from state I to S (-I, +S), exponential with rate duration_pf
  + fever - (-I, +F, +I), exponential distribution with rate mnFever, but its
    CDF f(infinity) = fever_pf. That is, there is a bernouilli probability of fever_pf
    that this fires.
  + treatment - (-I, +T) rate mnTreatPF
  + recover - (-I, +S). This is also called clearance.
* Chemoprophylaxis
  + treatment - seek treatment, (-I, +P)
  + chemoprophylaxis - duration of prophylaxis, (-P, +S), rate mnChemoprophylaxisPf
* Vaccination
  + PE - sporozooite-blocking vaccination, incomplete by U(0, 1) < pe_protect_pf.
    Lowers biting weight, b.
  + PE with treatment - same as PE, but it starts chemoprophylaxis.
  + GS - gametocyte-blocking vaccination. incomplete by U(0, 1) < gs_protect_pf. Lowers
    c, the probability to infect mosquitoes.
  + GS with treatment.
  + wane PE - (-PE, +S) distributed as normal(mnPE_PF, vr_PE_Pf). Sets b back to full power.
  + wane GS - (-GS, +S) distributed as normal(mnGE_PF, vr_GE_Pf). Sets c back to full power.
* Bites - These will move to another module, the biting module.
  + The biting algorithm returns a number of bites for this person.
  + The biting hazard is incomplete. There is a probability draw that
    U(0, 1) < b to create move the person to the latent state.


# Continuous Time Model

A continuous-time algorithm is a good choice when there will be few events in
each time step. In this case, most processes are slow relative to a day, so we
suspect continuous-time is a good choice, even in R where a single time step can
be expensive.

We can make a back-of-the-envelope calculation by assuming larger numbers for
each possible event. If each person gets ten bouts of
malaria in their lifetime, that's infection, treatment, and back to susceptible, for
thirty events. If we consider infectious bites, it's either those same infection events
or we're counting a lot more bites, in the thousands. For a ten year simulation, it's
3600 days in ten years. We either have events in 30/3600 days or over one per day.
In the former case, continuous time would likely be faster. In the latter case, discrete
time may be faster.

Given the many ways to implement continous-time models, our design will emphasize
a) getting the calculation correct and b) being able to make a new simulation by
changing parts of an old simulation. The design will take advantage of an important
constraint on the system, that the same model is applied across individuals.

We won't use the most practical design, where one event adds the next possible
events to a queue. For instance, an SEIR model might have an infection transition that
fires like this.
```{r usual_firing, eval = FALSE}
fire_infect <- function(state, individual) {
  state[individual, "disease"] <- "infected"
  add_to_queue(recovery_of(individual))
  add_to_queue(fever_of(individual))
  remove_from_queue(chemoprotection_of(individual))
}
```
That function gets the job done, but it combines the infection process with
assumptions about what other transitions are in the system. You can't define just
infection. You'd have to create a new infection transition for a system that doesn't
model fever.

Here's what we will do instead. Our continuous time algorithm will represent each
individual as a row in a data.table. We will find the next transition to fire by finding
the next individual whose soonest transition is next to fire. This is an old trick,
that the minimum time is the minimum of minima. This trick is helpful here because
the data.table class, as opposed to the base::data.frame class, has a feature that makes
it very fast to find the minimum of a sorted column.

If we can find the next individual to have a transition, then the code can look
just at firing times for this individual in order to determine which event is next.
At each time step, we ask every possible transition for the chosen individual whether
it would like to be enabled, or whether it is still enabled.
The algorithm, in pseudocode, is
```{r, eval = FALSE}
transition_to_fire <- which.min(individual_firing_times)
individual_state <- fire(transition_to_fire, individual_state)
disable(individual_firing_times, transition_to_fire)
previously_enabled <- enabled(individual_firing_times)
now_enabled <- apply(enabling_functions, individual_state)
update_firing_times(!previously_enabled & now_enabled)
disable(individual_firing_times, !now_enabled)
```
Compared to the algorithm that adds and removes from the queue, this algorithm
doesn't know, beforehand, which transitions to add or remove. It might be slower,
however, it is less likely we will forget to add or remove from the queue and we
can write transitions that are less tightly-coupled to one another. Moreover,
this style of update was shown, for chemical simulations, to be faster for small numbers
of transitions because simplicity of the code wins over algorithmic order.


## Implementation

We can represent the state above as a disease state, treatment state,
and scheduled bites.
```{r list_state}
individual <- list(
  disease = factor("S", levels = c("S", "I")),
  treatment = factor("U", levels = c("U", "PE", "GS")),
  bites = c(0.1, 0.4, 5.7, 7.4)
  )
```
Then we have thirteen transitions. We usually think about these rules
as, "when this happens, then that can happen," but we can, alternatively,
think about each transition as a separate entity that can happen whenever
the state allows it. In this case, each one is defined by

- is_enabled, a rule to determine from the new state whether the
  transition could happen.
- when, a function that reads the current state to draw a next time to fire.
  If the distribution is incomplete, this returns infinity for not firing.
- fire, a function that modifies the state when this fires.

This kind of definition decouples the transitions from each other.
It gets rid of questions about disabling from the queue, at the expense
of checking, at every time step, which is enabled.

We could store these as named lists.
```{r transition_list}
program_globals <- new.env()
program_globals$b <- 0.2
program_globals$c <- 3.7

is_enabled <- list(
  infect = function(state) with(state, disease == "S"),
  recover = function(state) with(state, disease == "I")
)
when <- list(
  infect = function(state) rexp(1, 1/50),
  recover = function(state) rexp(1, 1/200)
)
fire <- list(
  infect = function(state, time) {
    b <<- 0.3
    within(state, {disease = "I"})
    },
  recover = function(state, time) {
    b <<- 0.25
    within(state, {disease = "S"})
    }
)
for (set_env_idx in 1:length(fire)) {
  environment(fire[[set_env_idx]]) <- program_globals
}
```
For this individual, simulation is a loop that checks
every transitions enabling every time. The logic for this comes
from papers in chemical simulation that show that the first reaction
method is much faster than optimized direct or next reaction methods
when there are few transitions.
```{r simulate_individual}
# The current time and next firing times are part of the next state of the system.
# If a firing time is Inf, that means it isn't scheduled.
current_time <- 0
times <- vapply(
  when,
  FUN = function(x) x(individual),
  FUN.VALUE = vector(mode = "numeric", length = 1))

step_cnt <- 5
trajectory <- vector(mode = "list", length = step_cnt)
for (step_idx in 1:step_cnt) {
  to_fire <- which.min(times)
  if (is.infinite(to_fire)) break
  current_time <- times[[to_fire]]
  trajectory[[step_idx]] <- list(event = names(to_fire), when = current_time)
  individual <- fire[[to_fire]](individual, current_time)
  was_enabled <- is.finite(times)
  newly_enabled <- vapply(
    is_enabled,
    FUN = function(x) x(individual),
    FUN.VALUE = vector(mode = "logical", length = 1))
  times[was_enabled & !newly_enabled] <- Inf
  newly_enabled <- when[!was_enabled & newly_enabled]
  new_times <- current_time +
    vapply(
      newly_enabled,
      FUN = function(x) x(individual),
      FUN.VALUE = vector(mode = "numeric", length = 1))
  times[names(new_times)] <- new_times
}
do.call(rbind, trajectory)
```
```{r}
ls.str(program_globals)
```

That's all fun. So what I propose is to use a data.table because
it can be automatically sorted by time. Put the times of transitions into
each table row, along with individual state. The next-to-fire
will be the next-individual-to-fire.
```{r table_of_individuals}
individuals <- data.table(
  disease = factor(c("S", "I"), levels = c("S", "I")),
  treatment = factor(c("U", "U"), levels = c("U", "PE", "GS")),
  bites = list(c(0.1, 0.4, 5.7, 7.4), c(0.2, 0.3)),
  when = rep(Inf, 2),
  infect = rep(Inf, 2),
  recover = rep(Inf, 2)
  )
setorder(individuals, when)
```
Then we modify our algorithm above to use the first individual, because they
are sorted. The next steps follow as above.


# Discrete Time Model

We can think of a stochastic simulation as having a two parts,
how you update the state and how you sample what fires.
In order to update the state efficiently in R, it will help to represent
the state as a vector. R doesn't like loops. Let's use
a vector addition system (VAS).

## Update of the State
For a vector addition system, the state is a vector.
```{r human_state}
state <- c(S = 1, L = 0, I = 0, P = 0)
```

Then you create a set of transitions that reduce one part
of the vector and add to another. We can think of these as a matrix
where a -1 means you take from that state and a +1 means you add
to that state.
```{r human_transitions}
transitions <- matrix(
  c(
    -1, 1, 0, 0,
    0, -1, 1, 0,
    1, 0, -1, 0,
    1, 0, 0, -1,
    -1, 0, 0, 1,
    0, -1, 0, 1,
    0, 0, -1, 1
  ),
  nrow = length(state),
  dimnames = list(names(state), c("infect", "infectious", "recover", "unprotect",
                                  "protects", "protectl", "protecti"))
)
```
If we want to represent a self-transition, from protected to protected,
it would need to have a -1 and a 1 in the same matrix entry.
We can't represent it in this one matrix, but you'll see below
that we could represent it by using one matrix for the -1 and one for the 1.
A transition is enabled if it the state is nonzero
where there is a -1.
```{r find_enabled}
enable <- function(transitions) {
  transitions[transitions >= 0] <- 0
  -transitions
}
enable(transitions)
```
Then use that to find the enabled transitions.
```{r enabled_multiply}
enabled <- state %*% enable(transitions)
enabled
```
```{r show_enabled}
transitions[, enabled > 0]
```
Firing looks like, well, vector addition.
```{r fire_example2}
state + transitions[, enabled > 0]
```

## Sampling the probability distribution

Most of the distributions in the current PfSI are exponential distributions.
This version would be in discrete time, so the equivalent is a geometric
distribution. Sampling geometric distributions is particularly simple,
because you draw from the same binomial distribution for each time step.
Not all of the distributions are geometric, though, so we have to figure out
a more general strategy.

### Marginal vs Conditional

A discrete probability distribution is of the form $P[X=k]$, where the random
variable $X$ is the time step when the event occurs, counting from the time
step when the transition is enabled. We disallow $k=0$ because it confuses our
definition of what is one event or part of an event.

When we simulate from a discrete
probability distribution, we have two choices. We can draw $P[X=k]$ directly, as
the time, $t+x$, at which the event will happen. We can alternatively draw
$P[Y=t+x_i|i>=i-1]$ at each time step. This conditional sample asks, what is the probability
that the event happens now given that it has not yet happened? For a geometric
distribution, this is a sample from a Bernoulli distribution. For other distributions,
the conditional sample can be more difficult to calculate.

There is a little more information in the conditional sample because it tells
you the probability for firing at each time step. That extra information would
make it possible to generate a discrete hazard rate and look at likelihood.
We aren't doing that now.

The two ways to draw change how we define the state of the system.
If we draw $P[X=k]$ directly, then the state of the system is the state of
individuals and the time at which all enabled transitions will fire.
If we use a conditional sample, then the state of the system is the state
of individuals and the time at which each enabled transition became
enabled.


### Conflicts

What if one transition removes an individual but another recovers
that individual, in the same time step? These conflicts happen all of
the time in discrete simulation, and we decide each conflict either
by creating rules about who goes first or by calculating the probability
of either outcome, given the simulation's inputs.

A rule could say that, if an individual can recover or be removed, they
will always be removed. Every possible conflict in the system needs a
rule. One way to state those rules is to rank which transitions precede
each other. The earlier rule is the one that fires, and the later rule
does not get to fire for this particular time step.

The other way to resolve conflicts is to treat each transition probability
as a conditional statement. It's the probability someone would be removed
if there were no recovery. It's the possibility of recovery
if there were no removal. Given that the probability distributions come
from survival analysis on patient data, or some equivalent, each transition's
distribution is a conditional probability, so that's appropriate.
If you start from this assumption, write out the conditional probability,
and combine it with completeness relations, you can use some algebra
to derive the marginal probability of recovery or removal. In the end,
it becomes a coin flip between the two. This method is covered in [@Molloy1985-bk].

There is a quick fix to conflicts, which is to precalculate a probability
that, given two events in conflict, one fires before the other. Remember that
the conditional probability of firing on any one time step changes,
unless the distributions are exactly geometric, so that precalculation will,
in general, be an approximation. Still, it can offer a little better
result than using a simple precedence rule.

Keep in mind that the standard way to deal with conflicts is to reduce the
time step low enough that conflicts are rare.





# Bibliography
