---
title: "Mosquito Module Description"
abstract: "Summarizes two choices: the protocol for module interaction and how it runs across time steps."
header-includes:
  - \usepackage{amsmath}
output:
  pdf_document:
    toc: true
    df_print: kable
    number_sections: yes
  html_notebook:
    df_print: paged
---


```{r load_library, echo = FALSE}
library(macro)
```

# Introduction

A discrete-time simulation is a system of interacting transitions,
and we construct simulations, and customize them, by adding or removing
transitions, or by changing parameters. The MASH software offers
its users and developers two distinctly different ways to configure
a custom simulation. Our "modules" are the first way to configure
a simulation, and the second way is all the other ways you
change the behavior of R code. This could include class structures,
function arguments, strategy patterns, the works.

We have two different ways to configure simulations because
one of those ways, what we call "modules," is intended to allow
very wide latitude in how the module is implemented. Our
goal is to incorporate simulations from different research
groups, possibly in different computer languages. We can
do this by being very precise about how modules interact.

We only make a module when

a) This part of the simulation is the expertise of a
   different set of domain experts. We think those experts
   would like to write their own simulation, however
   they like, in order to plug it in.
   
b) We can promise that all reasonable versions of this
   module will receive and send data that is so
   similar that it uses the exact same data frame
   columns and rows, with the same meaning.

Otherwise, we aren't doing anybody any favors by calling
something a module. It's just R code that's more
difficult to write.


# Software Configuration

## Modules

The module diagram, which we are fond of drawing, shows
four boxes and directed arrows, but the diagram doesn't
explain what a box or arrow means for the code or for
the users and developers of that code.

```{r image-ref-for-in-text, echo = FALSE, message=FALSE, fig.align='center', fig.cap='MASH modules exchange data. The green lines cross discrete-time steps.', out.width='0.3\\linewidth'}
knitr::include_graphics("./cyclic_module_diagram_crop.pdf")
```

In the code, each of those boxes is an S3 class. We create
instances of the S3 classes in order to construct a simulation.
Take a look at a health module called `forced_si`.
You'll see that we pass bites in and retrieve the "path"
of the module over the time step, as its result.

```{r s3_module, eval = FALSE}
mash_step.forced_si <- function(module, bites) {...}
human_disease_path.forced_si <- function(module) {...}
```

The value of the `human_disease_path` function is a
data frame, like this example.

```{r health_status_a}
sample_health_infection_status()
```

The arrows in the diagram are a visual representation
of the main loop of the code. They represent each call
of a module and passing its output to the next module.
```{r main_loop, code = capture.output(dump("step_mainloop", ""))}
```

If you wanted to create another health module, you would
use this exact same output format. We make the claim that
this exact data format will be sufficient for a wide range
of simulations that probe the most important behaviors of
malaria. Defining these protocols is the most important
work we have done here because it requires judgement
that the largest, unsolved questions about malaria will
be addressed by individual-level human models, that
have human movement, that assign bloodmeals with biased
distributions, from spatial mosquito populations.

## Other variations in simulation

If you want to modify mosquito larval stages, there
isn't a separate module for mosquito larval stages.
You can take a mosquito module and create a variation of
it. For instance, you could pass a parameter to the
module that chooses a different strategy for birth
rates. We don't put limits on how you create points of
variation within a module.

Let's say you wanted to create a common simulation
framework for larval mosquitoes. You could make a function
```{r mosquito_larvae}
larval_step <- function(larval_state, adult_females) {...}
```
Then you could call it from any mosquito module you choose,
but that mosquito module has to accept this larval step
as a parameter. Its code has to cooperate. This is easier
than creating a separate larval module. The larval routine
can give and get whatever it wants, in any format.


# The lags of TWICE

As a reminder, we worked out that we would change the
time step between the output of the bloodmeal module
and the inputs to the health and mosquito modules.
This module diagram is incomplete because there is no
message from health status to location, and it reflects the current
set of data that are passed among the modules.

```{r image-ref-again, echo = FALSE, message=FALSE, fig.align='center', fig.cap='MASH modules exchange data. The green lines cross discrete-time steps.', out.width='0.3\\linewidth'}
knitr::include_graphics("./cyclic_module_diagram_crop.pdf")
```

The green arrows are where data from one time step is sent
to a module that will be computing for the next time step.
That means that the bloodmeal doesn't receive any bites that
happen during the week (or so) it is currently calculating. Similarly,
the human module doesn't receive any bites that happen during
the week it is currently calculating.

# Protocols

The current code sends data among modules in a format
described below. I give samples of that format and discuss
how I would like to change it.

## Health infection status

The current version of health infection status has one row per individual.
The `Start` column is their initial state of infection. Each additional pair
of `Time` and `Level` columns documents a change to that state. The state of
infection in the example is either a 0 or a 1, but it could be a float.
We haven't pinned that down. The columns are ID, Start, Time1, Level1,
Time2, Level2, and so on.
```{r health_status}
sample_health_infection_status()
```

I propose that we change this format from wide to long, which would
mean there are three columns, one for the `ID`, one for the `Time`,
and one for the `Level`. I propose this because, when I wrote the code
in the bloodmeal module to process this data, I thought it would have
been easier to write if the dataframe were long instead of wide.

## Location

The location protocol is similar to the health protocol. It's a wide
dataframe, as shown below.
```{r location_protocol}
sample_move_location()
```

I propose to make this a long dataframe, for the same reason, that
working with the wide dataframe was difficult, and using a long
version seemed like it would be easier.

## Mosquito to bloodmeal

This output from the mosquito module to the bloodmeal module
is a list of locations, times, and bite levels, where the bite level is 
a measure of infectiousness. This one was fine to work with.
```{r mosquito_protocol}
sample_mosquito_kappa()
```


## Bloodmeal output

There are two outputs from the bloodmeal, one to the human module,
one to the mosquito module. First, a sample of output to the
mosquito module shows the detailed version that I think could
be reduced to $\kappa$ for a location.

```{r mosquito_protocol_humans}
sample_mosquito_eip()
```


Output to the human module shows that the bloodmeal module
converts from bites to an EIP and from a $\kappa$ to bites.
The heart of this transformation is the distribution of bites
to humans.
```{r bloodmeal_to_human}
data.frame(Location=c(2,2,2), Bite=c(.53, .44, .62), Time=c(5.9, 6.21, 6.23),
           Level=c(1, 0, 1), ID=c(4, 9, 4))
```

