---
title: "Mosquito Module Description"
abstract: "Summarizes two choices: the protocol for module interaction and how it runs across time steps."
header-includes:
  - \usepackage{amsmath}
output:
  pdf_document:
    toc: true
    df_print: kable
    number_sections: yes
  html_notebook:
    df_print: paged
---


```{r load_library}
library(macro)
```

# Introduction

I'm almost done implementing a first version of MASH, and I'd like
to highlight two choices, and what I've learned about those
choices by writing the code. My goal is to ensure we understand
those choices, so we can fix them now, if necessary.

# The time step

As a reminder, we worked out that we would change the
time step between the output of the bloodmeal module
and the inputs to the health and mosquito modules.
This module diagram is incomplete because there is no
message from health status to location, and it reflects the current
set of data that are passed among the modules.

```{r image-ref-for-in-text, echo = FALSE, message=FALSE, fig.align='center', fig.cap='MASH modules exchange data. The green lines cross discrete-time steps.', out.width='0.3\\linewidth'}
knitr::include_graphics("./cyclic_module_diagram_crop.pdf")
```

The green arrows are where data from one time step is sent
to a module that will be computing for the next time step.
That means that the bloodmeal doesn't receive any bites that
happen during the week (or so) it is currently calculating. Similarly,
the human module doesn't receive any bites that happen during
the week it is currently calculating.

# Protocols

The current code sends data among modules in a format
described below. I give samples of that format and discuss
how I would like to change it.

## Health infection status

The current version of health infection status has one row per individual.
The `Start` column is their initial state of infection. Each additional pair
of `Time` and `Level` columns documents a change to that state. The state of
infection in the example is either a 0 or a 1, but it could be a float.
We haven't pinned that down. The columns are ID, Start, Time1, Level1,
Time2, Level2, and so on.
```{r health_status}
sample_health_infection_status()
```

I propose that we change this format from wide to long, which would
mean there are three columns, one for the `ID`, one for the `Time`,
and one for the `Level`. I propose this because, when I wrote the code
in the bloodmeal module to process this data, I thought it would have
been easier to write if the dataframe were long instead of wide.

## Location

The location protocol is similar to the health protocol. It's a wide
dataframe, as shown below.
```{r location_protocol}
sample_move_location()
```

I propose to make this a long dataframe, for the same reason, that
working with the wide dataframe was difficult, and using a long
version seemed like it would be easier.

## Mosquito to bloodmeal

This output from the mosquito module to the bloodmeal module
is a list of locations, times, and bite levels, where the bite level is 
a measure of infectiousness. This one was fine to work with.
```{r mosquito_protocol}
sample_mosquito_half_bites()
```

I have a question about this protocol. The first mosquito module is designed
to mirror the Ross-Macdonald model, and it depends only on a rate of infectiousness
from humans to mosquitoes and vice-versa. Do you anticipate that we could support
future work if we change this protocol to be that level, instead?

```{r maybe_mosquito}
data.frame(Location = c(1, 2, 3, 3, 1), Time = c(0.0, 0.0, 0.0, 1.0, 3.0),
           AttackRate = c(0.1, 0.12, 0.9, 0.2, 0.8))
```
Changing the protocol to an attack rate over time, that changes, would
require that we create individual bites from an inhomogeneous Poisson process
inside the bloodmeal module. We have to do that work, anyway, inside the
mosquito module right now, but it shifts to the bloodmeal module.

If we don't limit the protocol to the attack rate and, instead, keep the list
of individual bites, then I notice that the bites are identified by
location and time but not by mosquito. This protocol would mean that
the bloodmeal would lose information about which mosquito bit
a naive, or an infectious, human. It's not able to fully support an
individual-level mosquito simulation.

I'm suggesting we move what we have in one of two directions, more
granular or more detailed.

## Bloodmeal output

There are two outputs from the bloodmeal, one to the human module,
one to the mosquito module. First, a sample of output to the
mosquito module shows the detailed version that I think could
be reduced to $\kappa$ for a location.

```{r bloodmeal_output_mosquito}
data.frame(Location = c(2, 3), Bite = c(0, 0), Time = c(.27, 6.24), Level = c(1,1))
```

Then let's look at a sample of output to the human module.
```{r bloodmeal_to_human}
data.frame(Location=c(2,2,2), Bite=c(.53, .44, .62), Time=c(5.9, 6.21, 6.23),
           Level=c(1, 0, 1), ID=c(4, 9, 4))
```

# Conclusion

These are the questions:

1. Should we change protocols from wide to long?

2. Should we change mosquito module protocol from individual bites
   to attack rate and $\kappa$ at each location and time?
